---
title: "Homework 4"
author: "Jack Cunningham"
format: pdf
editor: visual
---

```{r}
library(astsa)
```

# 1.

3.5)

We are tasked with showing that $x_{t} -x_{t-1}=\delta+w_t+y_t-y_{t-1}$ is stationary and $w_t$ and independent of $y_t$.

This is assuming that the model is $x_{t}=\mu_t+y_t$, with $y_t$ a stationary process.

For a time series to be stationary it must have a constant mean that does not depend on $t$ and a covariance function that only depends on lag h.

We first look at the constant mean that does not depend on t:

$$
E[x_t-x_{t-1}]=E[\delta+w_t+y_{t}-y_{t-1}]=\delta+E[y_t]-E[y_{t-1}]
$$

Then, as $y_t,y_{t-1}$ are stationary processes their expected value would be constant and not depend on $t$.

Next we look at the covariance function.

$$
\gamma(0)=Var(x_t-x_{t-1})=Var(\delta+w_t+y_t-y_{t-1})=Var(w_t)+Var(y_t)+Var(y_{t-1})-2Cov(y_t,y_{t-1})
$$

Since $y_t,y_{t-1}$ are stationary processes their covariance depends on lag h. So:

$$
\gamma(0)=\sigma_w^2+2 \gamma_y(0)-2\gamma_y(1)
$$

$$
\gamma(1)=Cov(x_{t}-x_{t-1},x_{t+1}-x_t)=Cov(\delta+w_{t}+y_t-y_{t-1},\delta+w_{t+1}+y_{t+1}-y_{t})
$$

$$
=Cov(y_t,y_{t+1})-Cov(y_t,y_t)-Cov(y_{t-1},y_{t+1})+Cov(y_{t-1},y_t)=\gamma_y(1)-\gamma_y(0)-\gamma_y(2)+\gamma_y(1)=2\gamma_y(1)-\gamma_y(0)-\gamma_y(2)
$$

We can see that $\gamma(1)$ also depends only on lag h.

This pattern continues on, with $\gamma(h)$ depending only on linear combinations of $\gamma_y(h)$ and thus depending only on h.

With this we can say $x_t-x_{t-1}$ is a stationary process.

3.6)

```{r}
n = length(varve)
varve_1 = varve[1:(n/2)]
varve_2 = varve[(n/2):n]
var_1 = var(varve_1)
var_2 = var(varve_2)
```

The sample variance in the first half of the data is `r round(var_1,3)` , in the second half of the data it is `r round(var_2,3)`. The large difference shows evidence of non constant variable, it seems variance is getting larger as time passes.

A common technique to address non-constant variance is to apply a log transformation. This is attempted below:

```{r,echo=FALSE}
par(mfrow = c(2,1))
log_varve = log(varve)
log_varve_1 = log_varve[1:(n/2)]
log_varve_2 = log_varve[(n/2):n]
log_var_1 = var(log_varve_1)
log_var_2 = var(log_varve_2)
tsplot(varve, main = "varve", ylab = "", col = 4)
tsplot(log_varve, main = "log(varve)",ylab = "",col = 4)
```

In the original Varve data there is strong evidence of non-constant variance visually, it can easily be seen that early oscillations have much smaller magnitude than later ones.

The log transformation brings the oscillation magnitudes closer, approaching constant variance.

A log transformation can also assist the transformation to normality.

```{r,echo=FALSE, fig.height=10}
par(mfrow = c(2,1))
hist(varve, col = 4,breaks = 15)
hist(log_varve, col = 4, main = "Histogram of log(varve)", breaks = 15)
```

The Varve data exhibits a long right tail. The log transformed varve data is close to symmetrical, moving it closer to normality.

b\)

```{r}
tsplot(log_varve, col = 4,ylab = "",main = "log(varve)")
```

From time period 200 to 400 we can see a similar upward trend to the temperatures in figure 1.9.

```{r}
acf(log_varve)
```

The sample ACF shows evidence of prolonged positive correlation that trends to zero very slowly.

d\)

```{r}
diff_log_varve = diff(log_varve)
tsplot(diff_log_varve, main = "Differenced log(varve)",ylab = "",col = 4)
```

When examining the time series plot of the differenced log transformed varve we see fast choppy oscillations with no evidence of a trend or correlation. This plot looks similar to previous stationary series' that we have analyzed in the past.

```{r}
acf(diff_log_varve)
```

The sample ACF plot is consistent with the previous plot, the only lag with a significant auto-correlation is lag 1, after which the correlation quickly trends to zero consistent with the behavior we would expect in a stationary series.

The differenced log transformed varve data can be interpreted as the year over year percentage change in global varve thickness from Massachusetts . This is because of the below:

$$
\log(y_t)-\log(y_{t-1})=\log(y_t/y_{t-1})\approx(y_t-y_{t-1})/y_{t-1}
$$
