---
title: "STAT 626 Homework 2"
author: "Jack Cunningham"
format: pdf
editor: visual
---

```{r, echo=FALSE}
library(astsa)
```

# 1.

2.2)

We have $x_t=\beta_0+\beta_1t+w_t$ where $w_t$ is white noise, so mean of 0 and variance of $\sigma^2_w$.

a\)

In order for $x_t$ to be stationary we must have $E[x_t]$ constant that does not depend on $t$. And $Cov(x_t,x_s)$ only depends on the lag $|t-s|$ and not the actual values of t and s.

We start with the first condition.

$$
E[x_t]=\beta_0+\beta_1t
$$

This clearly depends on t, so $x_t$ is not stationary.

b\)

We have $y_t=x_t-x_{t-1}$ when we rewrite this we have:

$$
y_t=x_t-x_{t-1}=\beta_0+\beta_1 t+w_t-(\beta_0+\beta_1(t-1)+w_{t-1})=\beta_1+w_t-w_{t-1}
$$

We start with the first condition of stationary:

$$
E[y_t]=\beta_1
$$

Since this is constant and does not depend on t, this condition is met.

Next we need to show that $Cov(y_t,y_s)$ only depends on the lag $|t-s|$:

$$
Cov(y_t,y_s)=Cov(\beta_1+w_t-w_{t-1},\beta_1+w_s-w_{s-1})
$$

When $s = t$ we have:

$$
Cov(y_t,y_t)=Var(y_t)=Var(\beta_1+w_t-w_{t-1})=2\sigma_w^2
$$

When $s = |t -1|$ we have:

$$
Cov(y_t,y_{t-1})=Cov(\beta_1+w_t-w_{t-1},\beta_1+w_{t-1}-w_{t-2})=Cov(-w_{t-1},w_{t-1})=-Var(w_{t-1})=-\sigma_w^2
$$

When $s = |t-2|$ we have:

$$
Cov(y_t,y_{t-2})=Cov(\beta_1+w_t-w_{t-1},\beta_1+w_{t-2}-w_{t-3})=0
$$So we have:

$$
\gamma(h) = \begin{cases} 2 \sigma^2_w & h=0 \\ -\sigma_w^2 & h=1 \\ 0 & h \geq2\end{cases}
$$Since we have shown that the auto covariance function depends on $s$ and $t$ only through the lag $h$ we can say $y_t$ is stationary.

c\)

We the two sided moving average $v_t=\frac{1}{3}(x_{t-1}+x_t+x_{t+1})$. To start, we expand out each $x$:

$$
v_t=\frac{1}{3}(\beta_0+\beta_1(t-1)+w_{t-1}+\beta_0+\beta_1(t)+w_{t}+\beta_0+\beta_1(t+1)+w_{t+1})=\frac{1}{3}(3\beta_0+3\beta_1+w_{t-1}+w_t+w_{t+1})
$$

Then to get the mean we take the expected value:

$$
E[v_t]=E[\beta_0+\beta_1+w_{t-1}+w_t+w_{t+1}]=\beta_0+\beta_1
$$

2.3)

We have the two sided moving average smoother $x_t=\frac{1}{4}(w_{t-1}+2w_t+w_{t+1})$ with $w_t$ white noise:

To find the the autocovariance function we start with the case of $h=0$:

$$
\gamma(h)=Cov(x_t,x_t)=Var(\frac{1}{4}(w_{t-1}+2w_t+w_{t+1}))=\frac{1}{16}(\sigma_w^2+4 \sigma_w^2+\sigma_w^2)=\frac{6}{16}\sigma_w^2=\frac{3}{8}\sigma_w^2
$$

Then the case of $h=1$:

$$
\gamma(1)=Cov(x_{t+1},x_t)=Cov(\frac{1}{4}(w_t+2w_{t+1}+w_{t+2}),\frac{1}{4}(w_{t-1}+2w_t+w_{t+1}))=\frac{1}{16}(Cov(w_t,2w_t)+Cov(2w_{t+1},w_{t+1}))
$$

$$
=\frac{1}{16}(2\sigma_w^2+2 \sigma^2_w)=\frac{1}{4}\sigma^2_w
$$

Then the case of $h=2$:

$$
\gamma(2)=Cov(x_{t+2},x_t)=Cov(\frac{1}{4}(w_{t+1}+2w_{t+2}+w_{t+3}),\frac{1}{4}(w_{t-1}+2w_t+w_{t+1}))=\frac{1}{16}Cov(w_{t+1},w_{t+1})=\frac{1}{16}\sigma_w^2
$$

So we have the autocovariance function:

$$
\gamma(h) =\begin{cases} \frac{3}{8}\sigma_w^2 & h=0 \\
\frac{1}{4}\sigma_w^2 & h=1 \\ \frac{1}{16}\sigma_w^2 & h=2 \\ 0 & h \geq3\end{cases}
$$

Then we get the autocorrelation function by dividing each term by $\gamma(0)$ so:

$$
\rho(h)=\begin{cases} 1 & h=0 \\
\frac{2}{3}&h=1 \\
\frac{1}{6} & h =2 \\
0 & h \geq3
\end{cases} 
$$

Here is a sketch of the ACF:

```{r,echo=FALSE}
h = seq(-5,5)
rho = c(0,0,0,1/6,2/3,1,2/3,1/6,0,0,0)
plot(h,rho, main = "ACF")
```

2.4)

We have $x_t=\phi x_{t-1}+w_t$ where $w_t \sim wn(0,1)$ and $x_t$ is stationary and $x_{t-1}$ is uncorrelated with noise $w_t$.

a\)

To show this we use the fact that $x_t$ is stationary, and thus expected value is constant and does not depend on t. So:

$$
E[x_t]=E[\phi x_{t-1}+w_t]=\phi E[x_{t-1}]
$$

Since $x_t$ is stationary $E[x_t]=E[x_{t-1}]$, thus the only $E[x_t]$ that always has $E[x_t]=\phi E[x_{t-1}]$ is $E[x_t]=\mu_{xt}=0$.

b\)

$\gamma(0)=Var(x_t)=Var(\phi x_{t-1}+w_t)=\phi^2Var(x_{t-1})+1$

We use the fact that $x_t$ is stationary, and thus autocovariance only depends on the lag component and not t so say that in general $Var(x_i)=\gamma(0)$. Therefore $Var(x_{t-1})=\gamma(0)$:

$$
\gamma(0)=\phi^2\gamma(0)+1
$$

$$
\gamma(0)-\phi^2 \gamma(0)=1
$$

$$
\gamma(0)=\frac{1}{1-\phi^2}
$$

c\)

Since variance cannot be negative only $\phi <1$ makes sense.

d\)

First we want to find $\gamma(1)$:

$$
\gamma(1)=Cov(x_{t+1},x_t)=Cov(\phi x_{t}+w_{t+1},x_t)=Cov(\phi x_t,x_t)=\phi Var(x_t)=\phi \gamma(0)
$$

Then we can find the log-one autocorrelation easily:\

$$
\rho(1)=\frac{\gamma(1)}{\gamma(0)}=\frac{\phi \gamma(0)}{\gamma(0)}=\phi
$$

2.5)

We have $x_t=\delta + x_{t-1}+w_t$ with $x_0=0$ and $w_t \sim wn(0,\sigma^2_w)$.

a\)

To show this we start expanding $x_{t-1}$:

$$
x_t=\delta+\delta+x_{t-2}+w_{t-1}+w_t=\delta+\delta+\delta+x_{t-3}+w_{t-2}+w_{t-1}+w_t=\delta t+\sum_{k=1}^t w_k
$$

b\)

$$
E[x_t]=E[\delta t]+\sum_{k=1}^tw_k]=\delta t
$$

To find autocovariance I start with $\gamma(t,t)$:

$$
\gamma(0)=Var(x_t)=Var(\delta t+\sum_{k=1}^tw_k)=t \sigma^2_w
$$

Then $\gamma(t+1,t)$:

$$
\gamma(t+1,t)=Cov(x_{t+1},x_t)=Cov(\delta(t+1)+\sum_{k=1}^{t+1}w_k,\delta t +\sum_{k=1}^t w_k)=t \sigma^2_w
$$

The pattern becomes clear, the number of matching terms is what determines the coefficient in front of $\sigma_w^2$. Or in other terms:

$$
\gamma(s,t)=\min \{s,t \} \sigma_w^2
$$

c)

Stationary series have constant expected value that does not depend on t. $E[x_t]=\delta t$ explicitly depends on t.

d\)

$$
p_x(t-1,t)=\sqrt{\frac{t-1}{t}}, \lim_{t \rightarrow \inf}(\frac{t-1}{t})=1, \text{ so }\lim_{t \rightarrow \inf}\sqrt{\frac{t-1}{t}}=1
$$

This suggests that as we look farther and farther into the future, points that are near each other are related more and more.

e\)

I would use the transformation $y_t=x_t-\delta-x_{t-1}$. To prove this is stationary we start with its expected value:

$$
E[y_t]=E[x_t-\delta-x_{t-1}]=E[\delta+x_{t-1}+w_t-\delta-x_{t-1}]=E[w_t]=0
$$

Expected value is constant and does not depend on t.

Next we look at autocovariance, starting with $\gamma(0)$:

$$
\gamma(0)=Var(y_t)=Var(w_t)=\sigma_w^2
$$

Next $\gamma(1)$:

$$
\gamma(1)=Cov(y_{t+1},y_t)=Cov(x_{t+1}-\delta-x_t,x_t-\delta-x_{t-1})=Cov(w_{t+1},w_t)=0
$$

So we have the autocovariance function of:

$$
\gamma(h)=\begin{cases} \sigma_w^2 & h=0 \\ 0 & h \geq1\end{cases}
$$

Since autocovariance only depends on lag, this is a stationary series.

2.6)

I would treat this data as non stationary. The first necessary aspect of a stationary series is that expected value is constant and does not depend on t. There is an increasing trend in the data, as t increases the values of temperature deviations are increasing. So the expected value of $x_t$ depends on t.

2.9)

We have $x_t=w_tw_{t-1}$ with $w_t$ as white noise.

$$
E[x_t]=E[w_t w_{t-1}]=E[(w_t-0)(w_{t-1}-0)]=Cov(w_t,w_{t-1})=0
$$

Starting with $\gamma(t,t)$:

$$
\gamma(t,t)=Var(x_t)=Var(w_tw_{t-1})=E[(w_tw_{t-1}-\mu_{x_t})^2]=E[(w_tw_{t-1})^2]=Cov(w_t,w_{t-1})=0
$$

Then $\gamma(t+1,t)$:

$$
\gamma(t+1,t)=Cov(x_{t+1},x_t)=Cov(w_{t+1}w_t,w_tw_{t-1})=E[(w_{t+1}w_t)(w_t w_{t-1})]
$$

# 2.

We have $X_1$ with $Var(X_1)=1$ and $X_2$ with $Var(X_2)=4$, and $Cov(X_1,X_2)=-2$.

a\)

$Var(X_1-5X_2)=Var(X_1)+25Var(X_2)-2(5Cov(X_1,X_2))=1+(25)(4)-(2)(5)(-2)=121$

$Var(X_1+5X_2)=Var(X_1)+25Var(X_2) + 2(5)(Cov(X_1,X_2))=1+(25)(4)+(2)(5)(-2)=81$

$Var(X_1+X_2)=Var(X_1)+Var(X_2)+2Cov(X_1,X_2)=1+4+(2)(-2)=1$

b\)

$Var(X_1-cX_2)=Var(X_1)+c^2Var(X_2)-2c Cov(X_1,X_2)$

Take partial derivative relative to c:

$$
\frac{\delta}{\delta c}=2cVar(X_2)-2Cov(X_1,X_2)
$$

Set equal to zero:

$$
2cVar(X_2)=2Cov(X_1,X_2)
$$

$$
c=Cov(X_1,X_2)/Var(X_2)=\frac{-2}{4}=-1/2
$$
