---
title: "Homework 3"
author: "Jack Cunningham"
format: pdf
editor: visual
---

```{r}
library(astsa)
```

# I.

2.10)

We have $x_{t}=\mu+w_t+\theta w_{t-1}, \quad w_t \sim wn(0,\sigma^2_w)$.

a\)

$$
E[x_t]=E[\mu+w_t+\theta w_{t-1}]=\mu+0+0=\mu
$$

b\)

$$
\gamma_x(0)=Var(x_t)=Var(\mu+w_t+\theta w_{t-1})=Var(w_t)+\theta^2Var(w_{t-1})=\sigma_w^2(1+\theta^2)
$$

$$
\gamma_x(1)=Cov(x_{t+1},x_t)=Cov(\mu+w_{t+1}+\theta w_{t},\mu+w_t+\theta w_{t-1})=Cov(\theta w_t,w_t)=\sigma_w^2 \theta
$$

$$
\gamma_x(-1)=Cov(x_{t-1},x_t)=Cov(\mu+w_{t-1}+\theta w_{t-2},\mu + w+\theta w_{t-1})=Cov(w_{t-1},\theta w_{t-1})=\sigma^2_w \theta 
$$

c\)

For $x_t$ to be stationary expected value needs to be constant and not depend on t. The Auto-Covariance function needs to depend only on the lag $h=|s-t|$ and not the time $s$ or $t$. We can see both of these are met in the last two results. For $x_t$ to be stationary for all $\theta \in \mathbb{R}$ we simply need to show that $\gamma_x (0)=Var(x)$ is always non negative. Since $(1+\theta^2)$ is always non negative, we can say that $x_t$ is stationary for all $\theta \in \mathbb{R}$.

d\)

We have $Var(\bar{x})=\gamma_x(0)/n$ since $x_t$ is white noise.

i\)

$$
Var(\bar{x})=\gamma_x(0)/n=2 \sigma_w^2/n
$$

ii\)

$$
Var(\bar{x})=\gamma_x(0)/n=\sigma_w^2/n
$$

iii)

$$
Var(\bar{x})=\gamma_x(0)/n=2 \sigma_w^2/n
$$

e\)

The accuracy in the estimate of $\mu$ is best when $\theta = 0$, the variance of our estimate is 1/2 that of $\theta = 1 \text{ and } \theta = -1$.

2.11)

a\)

We are generating the series $x_t=w_t$ where $w_t \sim wn(0,\sigma_w^2)$.

```{r}
set.seed(3)
w = rnorm(500)
acf(w, lag.max = 20, plot = FALSE)
```

We find the actual ACF:

$$
\gamma_x(0)=Var(x_t)=\sigma_w^2
$$

$$
\gamma_x(1)=Cov(x_t,x_{t-1})=Cov(w_t,w_{t-1})=0
$$

So:

$$
\gamma_x(h) = \begin{cases} \sigma_w^2 & h = 0 \\ 0 &h \geq1\end{cases}
$$

And then the actual ACF is:

$$
\rho(h)= \begin{cases} 1 & h = 0 \\ 0 & h \geq1\end{cases}
$$

We can see that the sample ACF varies slightly around the true value 0.

b\)

```{r}
set.seed(4)
w_2 = rnorm(50)
acf(w_2, lag.max = 20, plot = FALSE)
```

Now the estimates of the ACF vary more than with $n = 500$.

2.12)

We are generating the series $v_t=\frac{1}{3}(w_{t-1}+w_t+w_{t+1})$ where $w_t \sim wn(0,\sigma_w^2)$.

```{r}
set.seed(11)
w = rnorm(500+51)
v = filter(w, sides = 2, filter = rep(1/3, 3))[100:500]
acf(v, lag.max = 20, plot = TRUE)
acf(v, lag.max = 20, plot = FALSE)
```

The actual ACF is as follows:

$$
\gamma_v(0)=Var(v_t)=\frac{1}{3}\sigma_w^2
$$

$$
\gamma_v(1)=Cov(v_{t+1},v_t)= Cov(\frac{1}{3}(w_t+w_{t+1}+w_{t+2}),\frac{1}{3}(w_{t-1}+w_t+w_{t+1})=\frac{2}{9} \sigma_w^2
$$

$$
\gamma_v(2)=Cov(v_{t+2},v_t)= Cov(\frac{1}{3}(w_{t+1}+w_{t+2}+w_{t+3}),\frac{1}{3}(w_{t-1}+w_t+w_{t+1})=\frac{1}{9} \sigma_w^2
$$

So we have:

$$
\gamma_v(h)= \begin{cases} \frac{3}{9} \sigma_w^2 & h = 0 \\
\frac{2}{9} \sigma_w^2 & h = 1 \\ \frac{1}{9}\sigma_w^2 & h = 2 \\ 0 & h \geq 3\end{cases}
$$

And the ACF is as follows:

$$
\rho_v(h) = \begin{cases} 1 & h = 0 \\ \frac{2}{3} & h = 1 \\ \frac{1}{3} & h=2 \\ 0 & h \geq 3\\ \end{cases}
$$

The sample ACF is pretty close to the true values.

b\)

```{r}
set.seed(15)
w_2 = rnorm(50+51)
v_2 = filter(w, sides = 2, filter = rep(1/3, 3))[2:51]
acf(v_2, lag.max = 20, plot = TRUE)
acf(v_2, lag.max = 20, plot = FALSE)
```

We can see that the estimates start to vary from the true values by quite a bit, lag 2 is .169 which is quite far from the true 1/3, and lags 3 and 4 are both far from 0. By reducing n we have increased the variance in our estimator.

2.13)

```{r}
set.seed(1000)
w = rnorm(250 + 50)
x = filter(w, filter = c(1.5, -.75), method = "recursive")[-(1:50)]
acf(x, lag.max = 50, plot = TRUE)
```

We can see a strong negative correlated cycle starting at lag 4, that alternates to positive correlation and back to negative correlation about every 8 observations. The structure gradually fades to near zero as observations become more distant.

2.14)

```{r, fig.height=25, fig.width=10}
t = 1:500
cs = 2*cos(2*pi*(t+15)/50)
w = rnorm(500)
par(mfrow = c(3,1))
acf(cs,lag.max = 100, plot = TRUE)
acf(cs + w,lag.max = 100, plot = TRUE)
acf(cs + 5*w,lag.max = 100, plot = TRUE)
```

We can see that as the white noise component becomes larger the magnitude of the cycle becomes smaller in the ACF.
