---
title: "Homework 5"
author: "Jack Cunningham"
format: pdf
editor: visual
---

# I.

```{r, loading packages}
library(astsa)
```

4.4)

```{r,echo=FALSE}
par(mfcol = c(2,3))
ACF_1_1 = ARMAacf(ar = .6, ma = .9, 24)[-1]
PACF_1_1 = ARMAacf(ar = .6, ma = .9, 24, pacf = TRUE)
ACF_1_0 = ARMAacf(ar = .6, ma = 0, 24)[-1]
PACF_1_0 = ARMAacf(ar = .6, ma = 0, 24, pacf = TRUE)
ACF_0_1 = ARMAacf(ar = 0, ma = .9, 24)[-1]
PACF_0_1 = ARMAacf(ar = 0, ma = .9, 24, pacf = TRUE)

tsplot(ACF_1_1, type="h", xlab="lag", ylim=c(-.5,1),main = "ARMA(1,1)")
abline(h=0)
tsplot(PACF_1_1, type="h", xlab="lag", ylim=c(-.5,1))
abline(h=0)

tsplot(ACF_1_0, type="h", xlab="lag", ylim=c(-.5,1), main = "ARMA(1,0)")
abline(h=0)
tsplot(PACF_1_0, type="h", xlab="lag", ylim=c(-.5,1))
abline(h=0)

tsplot(ACF_0_1, type="h", xlab="lag", ylim=c(-.5,1), main = "ARMA(0,1)")
abline(h=0)
tsplot(PACF_0_1, type="h", xlab="lag", ylim=c(-.5,1))
abline(h=0)
```

The ARMA(1,1) model has an ACF plot that tends toward 0 over time and a PACF that alternates positive and negative that tends toward zero.

The ARMA(1,0) model has a similar ACF plot as ARMA(1,1), tending to 0 over time but with a smaller magnitude at all lags. The PACF that has correlation at lag 1 but none at later lags.

The ARMA(0,1) model has a ACF plot that has a positive value at lag 1 but none at further lags. the PACF has alternating positive and negative correlations that tend toward zero but at a lower magnitude when compared to ARMA(1,1).

Recognizing these patterns would allow us to choose the correct model from sample data.

```{r,echo=FALSE, fig.height=8,fig.width=10}
par(mfcol = c(2,3))
ARMA_1_1_sim = arima.sim(list(order = c(1,0,1), ar = .6, ma = .9), n = 100)
ARMA_1_0_sim = arima.sim(list(order = c(1,0,0), ar = .6), n = 100)
ARMA_0_1_sim = arima.sim(list(order = c(0,0,1), ma = .9), n = 100)


acf(ARMA_1_1_sim, main = "ARMA(1,1)")
pacf(ARMA_1_1_sim)

acf(ARMA_1_0_sim, main = "ARMA(1,0)")
pacf(ARMA_1_0_sim)

acf(ARMA_0_1_sim, main = "ARMA(0,1)")
pacf(ARMA_0_1_sim)
```

The sample ACF and PACFs are pretty similar to the theoretical values, there are a few cases where there are lags that are significant that we wouldn't expect. In particular ARMA(1,0) has lags that are significant in the negative direction in lag 15 that we wouldn't expect as it should tend to zero. The same idea can be seen in AR(0,1) we expect all those lags to be zero past lag 1 but we see significance in lag 17.

```{r,echo=FALSE, fig.height=8,fig.width=10}
par(mfcol = c(2,3))
ARMA_1_1_sim = arima.sim(list(order = c(1,0,1), ar = .6, ma = .9), n = 500)
ARMA_1_0_sim = arima.sim(list(order = c(1,0,0), ar = .6), n = 500)
ARMA_0_1_sim = arima.sim(list(order = c(0,0,1), ma = .9), n = 500)


acf(ARMA_1_1_sim, main = "ARMA(1,1)")
pacf(ARMA_1_1_sim)

acf(ARMA_1_0_sim, main = "ARMA(1,0)")
pacf(ARMA_1_0_sim)

acf(ARMA_0_1_sim, main = "ARMA(0,1)")
pacf(ARMA_0_1_sim)
```

With n = 500 the patterns are closer to the theoretical values but we see more correlations poking into significance, likely due to the increased power from increasing n.

4.5)

a\)

```{r}
tsplot(diff(cmort), main = "Differenced Mortality")
```

The original mortality data appears to have a small downward trend, looking like a random walk with drift.

The differenced mortality data looks like white noise.

Differencing seems reasonable because when we are dealing with a random walk with drift situation the operation yields a stationary process.

b\)

```{r, fig.width=10}
par(mfrow = c(1,2))
acf(diff(cmort), main = "ACF Differenced Mortality")
pacf(diff(cmort), main = "PACF Differenced Mortality")
```

The ACF starts large and then tails to zero over time. The PACF at lag one is about -.5 and then cuts off.

Using table 4.1, we can see that AR(1) makes sense for this data as the ACF tails off and the PACF cuts off after the first lag.

c\)

```{r,echo=FALSE, output = FALSE, message=FALSE}
model <- sarima(diff(cmort), p = 1, d = 0, q = 0, no.constant = TRUE)
```

```{r,echo=FALSE}
model$fit
```

The AR1 coefficient is clearly significant. The white noise variance is estimated at 33.81.

d\)

```{r,echo=FALSE,fig.width=8, fig.height=10}
par(mfrow = c(2,2))
acf(resid(model$fit), main = "Residual ACF")
qqnorm(resid(model$fit))
qqline(resid(model$fit), main = "Residuals Normal QQ Plot")
tsplot(resid(model$fit), main = "Residuals")
```

We can see the residual ACF plot doesn't show any significant lags, the residuals appear normally distributed and the residual time series plot looks like white noise.

e\)

```{r,echo=FALSE}
sarima.for(diff(cmort), n.ahead = 4, p = 1, d = 0, q = 0)
```

f\)

The forecast and prediction interval is calculated through the m-step ahead forecast formula seen in formula 4.37. That is:

$$
x_{n+m}^n=\phi^{m}x_n
$$

This has the prediction error seen in formula 4.38:

$$
P_{n+m}^n=\sigma_w^2(1+\phi^2+ \dots+ \phi^{2(m-1)})
$$

g\)

Using the SARIMA forecast function with (p = 0, d = 1, q = 1) we can see the forecast is 85.46562.

```{r}
sarima.for(cmort, n.ahead = 1, p = 0, d = 1, q = 1)
```

4.6)

With an AR(1) model we have:\

$$
x_n=\phi x_{n-1}+w_n
$$

The step ahead forecast is the expected value of $E[x_{n+m}]$. We start with $m = 1$:

$$
E[x_{n+1}]=E[\phi x_{n}+w_{n+1}]=\phi x_n=x_{n+1}^n
$$

Then with $m = 2$:

$$
E[x_{n+2}]=E[\phi x_{n+1} +w_{n+2}]=\phi E[x_{n+1}]
$$

And we know that $E[x_{n+1}]=\phi x_n=x_{n+1}^n$ so:

$$
E[x_{n+2}]=\phi^2 x_n=x_{n+2}^n
$$

And we can say that generally we have:

$$
x_{n+m}^n=E[x_{n+m}]=\phi^m x_n
$$

Then we are tasked to show:

$$
E[(x_{n+m}-x_{n+m}^n)^2]=\sigma_w^2 \frac{1-\phi^{2m}}{1-\phi^2}
$$

We can write $x_{n+m}$ as:

$$
x_{n+m}=\phi^{m}x_n+\sum_{i=0}^{m-1} \phi^{i} w_{n+m-i}
$$

Which we can substitute into $E[(x_{n+m}-x^{n}_{n+m})^2]$:

$$
E[(x_{n+m}-x^n_{n+m})^2]=E[(\phi^mx_n +\sum_{i=0}^{m-1} \phi^i w_{n+m-i}-\phi^m x_n)^2]=E[(\sum_{i=0}^{m-1} \phi^i w_{n+m-i})^2]
$$

Since $E[\sum_{i=0}^{m-1} \phi^i w_{n+m-i}]=0$, we just need to find the variance of $\sum_{i=0}^{m-1} \phi^i w_{n+m-i}$:

$$
Var(\sum_{i=0}^{m-1} \phi^i w_{n+m-i})=\sum_{i=0}^{m-1} \phi^{2i} \sigma^2_w
$$

Which is a geometric series, so it equals

$$
=\sigma_w^2 (\frac{1-\phi^{2m}}{1-\phi^2})
$$

5.11)

We start our model process by investigating the time series plot:

```{r,echo=FALSE}
tsplot(birth)
```

This plot looks a lot like a random walk process. Variance appears constant so no issues there. It looks like there are seasonal effects.

I first try a difference to deal with the random walk process.

```{r,echo=FALSE}
tsplot(diff(birth))
```

This looks a lot closer to stationarity. To assess seasonality and choose p,q and P,Q we look at the ACF and PACF of the differenced data.

```{r,echo=FALSE}
par(mfrow = c(1,2))
acf(diff(birth),lag.max = 60)
pacf(diff(birth), lag.max = 60)
```

We see significant correlation between values a year apart. To go after this we different again using $x_{t-12}$:

```{r,,echo=FALSE}
tsplot(diff(diff(birth),12))
```

```{r,echo=FALSE}
par(mfrow = c(1,2))
acf(diff(diff(birth),12),lag.max = 60)
pacf(diff(diff(birth),12), lag.max = 60)
```

Now consulting the 5.1 table, we see that the ACF cuts off after the first 12 month lag and the PACF seems to tail off in 12 month intervals.

This would seem to indicate a $P = 0, Q = 1$ for the seasonal component of the $ARIMA(p,d,q) \times (P,D,Q)_s$.

In the non-seasonal component there looks like evidence for $q = 1$ as the ACF cuts off after the first lag. There is also evidence of $p = 4$ as the PACF pokes out out the first four lags then cuts very close to zero.

So the two candidate models I am considering are:

$$
ARIMA(0,1,1) \times (0,1,1)_{12}
$$ and:

$$
ARIMA(4,1,0) \times (0,1,1)_{12}
$$

The simpler model is the first one, so I start there:

```{r, fig.height=10,fig.width=8,echo=FALSE}
sarima(birth,p = 0, d = 1, q = 1, P = 0, D = 1, Q = 1,S = 12)
```

The $ARIMA(0,1,1) \times (0,1,1)_{12}$ model's biggest weakness is that there is still autocorrelation remaining in the residuals. We see this most clearly in the Ljung - Box statistic where the null hypothesis of white noise is rejected at each lag.

Because of this I consider the other model $ARIMA(4,1,0) \times (0,1,1)_{12}$:

```{r,fig.height=10,fig.width=8,echo=FALSE}
sarima(birth,p = 4, d = 1, q = 0, P = 0, D = 1, Q = 1,S = 12)
```

This model looks a lot better. Variance looks constant, the ACF plot and Ljung- Box statistic plots support the assumption of white noise errors and residuals look normal in the qq-plot.

Despite the diagnostics looking better, I am still concerned about the complexity of using p = 4. So I also try $ARMA(1,1,1) \times(0,1,1)_{12}$

```{r, fig.height=10, fig.width=8,echo=FALSE}
sarima(birth,p = 1, d = 1, q = 1, P = 0, D = 1, Q = 1,S = 12)
```

The models are comparable except for the p-value for Ljung-Box statistic plot, this model doesn't reject the null hypothesis around lag 7. But generally, the residuals are acceptable.

To make a final decision I compare the AIC/BIC of the two models:

| Model                              | AIC   | BIC   |
|------------------------------------|-------|-------|
| $ARIMA(0,1,4) \times(0,1,1)_{12}$  | 6.715 | 6.778 |
| $ARIMA(1,1,1) \times (0,1,1)_{12}$ | 6.722 | 6.765 |

Using AIC we would select the more complex model, using BIC we would select the simpler model. I choose $ARIMA(1,1,1) \times (0,1,1)_{12}$ due to it being a simpler model.

The 12 month forecast is below:

```{r,echo=FALSE}
sarima.for(birth,p = 1, d = 1, q = 1, P = 0, D = 1, Q = 1,S = 12,n.ahead = 12)
```

5.13)

a\)

We start by fitting an ARIMA model to the monthly sales data.

To begin, we look at the time series plot:

```{r,echo=FALSE}
tsplot(sales)
```

This looks like a combination of random walk and overall upward trend. To remedy this we difference:

```{r,echo=FALSE}
tsplot(diff(sales))
```

Variation appears constant, and this is a relatively smooth plot as nearby values look positively correlated. I can still see some evidence of upward trend from times 50 to 80 but for now I will look at the ACF and PACF of the data:

```{r,echo=FALSE}
par(mfrow = c(1,2))
acf(diff(sales))
pacf(diff(sales))
```

The ACF and PACF both seem to tail off, so my first model consideration is $ARIMA(1,1,1)$:

```{r, fig.height=10, fig.width= 8,echo=FALSE}
sarima(sales, p = 1, d = 1, q = 1)
```

Variance looks constant over time, the ACF and Ljung-Box statistic plot support the assumption of white noise residuals and the residuals look normally distributed. I am satisfied with the $ARIMA(1,1,1)$ model.

b\)

We start to investigate the relationship betweeen $\nabla S_t, \nabla L_t$ by using the CCF plot:

```{r,echo=FALSE}
ccf2(diff(sales), diff(lead))
```

The CCF plot shows significant correlation at lags 2 and 3, lets take at the lag plot for the earlier lags.

```{r,fig.width=10, fig.height=12,echo=FALSE}
par(mfcol = c(3,2))
lag2.plot(diff(lead), diff(sales), 5)
```

In this plot we can see the nature of the relationship between the two series. At lag 2, despite having a correlation of -.38, the relationship does not look linear. At lag 3 however the relationship is highly linear and has a strong positive correlation of 0.72. For this reason we should consider adding this to our model.

c\)

We are asked to fit $\nabla S_t=\beta_0+\beta_1 \nabla L_{t-3}+x_t$. To do this we first fit the linear model, $\nabla S_t=\beta_0+\beta_1 \nabla L_{t-3}+\epsilon$ .

```{r,echo=FALSE}
library(zoo)
pp = as.zoo(ts.intersect(diff_sales = diff(sales), diff_leadL3 = lag(diff(lead),-3)))
summary(reg <- lm(pp$diff_sales ~ pp$diff_leadL3))
```

We can take a quick look at the diagnostic for this simple linear regression:

```{r, fig.height = 8, fig.width = 8,echo=FALSE}
par(mfrow = c(2,2))
plot(reg)
```

The only concern would be some evidence of non-constant variance at the right end of the data. Other than that the model is well calibrated.

We take the residuals of that model and use it to fit the ARMA $x_t$ model.

```{r,echo=FALSE}
par(mfrow = c(1,2))
acf2(resid(reg))
```

The ACF tails off while the PACF cuts to zero after lag one. Using table 4.1 this indicates that we should use $ARMA(1,0)$.

```{r, fig.height=10, fig.width=8,echo=FALSE}
sarima(resid(reg), p = 1, d = 0, q = 0)
```

There is no pattern in the standardized residuals, variance looks steady, residuals look normally distributed and the ACF/Ljung-Box plots both indicate the residuals are white noise. This model fits very well.

We can compare the two models using AIC/BIC:

| Model                 | AIC   | BIC   |
|-----------------------|-------|-------|
| ARMA(1,1)             | 3.455 | 3.536 |
| AR(1) with Lead Lag 3 | 2.443 | 2.505 |

It is clear that the AR(1) model with the leading indicator is the superior model.

# II.

a\)

```{r,echo=FALSE}
set.seed(101)
w_t_1 = rnorm(150)
w_t_2 = rnorm(150)
w_t   = rnorm(100)

x_t_1 = filter(w_t_1, filter = c(.5), method = "recursive")[(-(1:50))]
x_t_2 = filter(w_t_2, filter = c(rep(0,11),.9), method = "recursive")[(-(1:50))]

y_t_1 = x_t_1 + cumsum(w_t)
y_t_2 = x_t_2 + 5*cumsum(w_t)
z_t   = 5*y_t_1 - y_t_2
```

Plotting:

```{r, fig.height = 12,echo=FALSE}
par(mfrow = c(3,1))
tsplot(y_t_1, main = "yt1")
tsplot(y_t_2, main = "yt2")
tsplot(z_t,   main = "zt")
```

The plots are very smooth which makes me think of random walk and cyclic time series plots, causing me to doubt they are stationary.

b\)

For a series to be stationary expected value must be constant and not depend on t and the auto-covariance function must depend only on lag h.

$$
E[y_{t_1}]=E[x_{t1}+\sum_{j=1}^t w_j]=E[x_{t1}]=E[0.5 x_{t-1,1} +w_{t1}]=0
$$

We start with lag = 0 for the auto-covariance function:

$$
Cov(y_{t1},y_{t1})=Var(y_{t_1})=Var(x_{t1}+\sum_{j=1}^t w_j)=Var(x_{t1})+t \sigma_w^2=\frac{1}{1-.5^2}+t
$$

Then lag = 1:

$$
Cov(y_{t+1,1,},y_{t,1})=Cov(x_{t+1,1}+\sum_{j=1}^{t+1}w_j,x_t+\sum_{j=1}^{t}w_j)=Cov(x_{t+1},x_t)+Cov(\sum_{j=1}^{t+1} w_j,\sum_{j=1}^t w_j)=\gamma(1)_{x1}+t=.5\frac{1}{1-.5^2}+t
$$

And generally:

$$
Cov(y_{t+h,1},y_{t,1})=\phi^h \gamma_{x_1}(0)+t
$$

Where $\gamma_{x_1}(0)=\frac{1}{1-\phi^2}$ and $\phi=.5$.

So $y_{t1}$ is not stationary because its autocovariance function depends on t.

c\)

We have $z_t=5y_{t1}-y_{t2}$. Starting with lag = 0:

$$
Cov(z_t,z_t)=Var(z_t)=Var(5y_{t_1}-y_{t2})=Var(5(x_{t1}+\sum_{j=1}^tw_j)-(x_{t2}+5 \sum_{j=1}^t w_j))=25Var(x_{t1})+Var(x_{t2})
$$

$$
=25 \frac{1}{1-\phi_{x_1}^2}+\frac{1}{1-\phi^2_{x_2}}
$$

Where, for ease of notation $\phi_{x_1}=.5$ and $\phi_{x_{2}}=.9$.

For lag=1:

$$
Cov(z_{t+1},z_t)=Cov(5y_{t+1,1}-y_{t+1,2},5y_{t,1}-y_{t,2})=Cov(5x_{t+1,t}-x_{t+1,2},5x_{t,1}-x_{t,2})=25\gamma(1)_{x1}+\gamma(1)x_{2}
$$

$$
=25 \phi_{x_1} \frac{1}{1-\phi^2 _{x_1}}
$$

So the autocovariance function is:

$$
\gamma(h)=25 \gamma(h)_{x1}+\gamma(h)_{x2}
$$

Where:

$$
\gamma(h)_{x_1}=\begin{cases} \frac{1}{1-.5^2} & h=0\\\
.5^h \gamma(0)_{x_1} & h\neq0\end{cases}
$$

and:

$$
\gamma(h)_{x_2}=.\begin{cases} \frac{1}{1-.9^2} & h=0\\
.9^{k}\gamma(0)_{x_2} & h = 12k,k=1,2,.. \\ 
0 & \text{otherwise} \end{cases}
$$

Since $E[z_t]=E[5y_{t1}-y_{t2}]=0$ , so constant and not depending on t and the auto-covariance function does not depend on t and only on the lag h we can say $z_t$ is a stationary process.

d\)

In part b we found that the reason why $y_{t1}$ was non-stationary was due to the sum of the white noise. in part c we found that $z_t$ was able to become stationary because it got rid of the sum of white noise leaving us with the stationary aspects of $y_{t1}$ and $y_{t2}$.

e\)

We start with lag 0 for the cross-covariance function:

$$
Cov(y_{t1},y_{t2})=Cov(x_{t1}+\sum_{j=1}^tw_j,x_{t2}+5\sum_{j=1}^t w_{j})=5t
$$

Then with lag 1:

$$
Cov(y_{t+1,1},y_{t,2})=Cov(x_{t+1,1}+\sum_{j=1}^{t+1}w_{j},x_{t2}+5 \sum_{j=1}^t w_j)=5t
$$

So when lag is greater than or equal to 0, the covariance is $5t$:

But when lag is less than zero it is a bit more interesting:

$$
Cov(y_{t-1,1},y_{t,2})=Cov(\sum_{j=1}^{t-1} w_j,5 \sum_{j=1}^tw_j)=5(t-1)
$$ So the cross-covariance function is:

$$
Cov(y_{t+h,1},y_{t,2})=\begin{cases}5t &h\geq0 \\ 5(t-h)& h <0  \end{cases}
$$

The cross-correlation function is:

$$
\rho_{xy}(y_{t+h,1},y_{t,2})=\frac{Cov(y_{t+h,1},y_{t,2})}{\sqrt{\gamma_{y1}(0) \gamma_{y_2}(0)}}
$$

The denominator would be:

$$
\sqrt{(\frac{1}{1-.5^2}+t)(\frac{1}{1-.9^2}+t)}
$$

The two time series are not jointly stationary because their cross-covariance function depends on t.
